%  This is the driver file for the CMB-S4 Science Book.

%%% it has been adapted from the Snowmass reports from the Cosmic Frontier  %%%%
%%% Working Groups to use bibtex. 19Sep15, T. Crawford (orig. 5/13/13 J. Feng) %%%%

%  D. Hitlin   9/23/03   derived from the BABAR Physics Book format

%  To use LATEX with this format, you must have the follwing files 
%  in the same directory as your text source and figure files
%  tcibook.cls
%  fancyhea.sty
%  work.sty
%  epsfig.sty
%  workshopsym.tex       This file provides macros for many common symbols
%                         Using these macros will provide uniformity of notation
%                         for the basic particle symbols, units, etc.
%
%  These provide the page size, type style, headings, etc.


\documentclass[titlepage]{tcibook}
\usepackage{fancyhea}
\usepackage{work}
\usepackage{bm}       %    enables bold math symbols  e.g.  \bm{\gamma}
\usepackage{graphicx}
\usepackage{hyperref}      % hypertext links %%ARXIV
\usepackage[usenames,dvipsnames]{color} %used for font color
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{chngcntr}
\usepackage{booktabs}
\usepackage{multirow}
\counterwithout{figure}{chapter}
%\usepackage[explicit]{titlesec}
\usepackage{titlesec}

\usepackage[utf8]{inputenc}

\input workshopsymbols.tex      %   standard macros for common HEP terms
\newcommand{\cmbexp}{{CMB-S4}}
%    
% Abbreviations for ADS bibtex entries
%
\newcommand\mnras{{\it Mon. Not. Roy. Astron. Soc.}}
\newcommand\apj{{\it Ap.\ J.\ }}
\newcommand\apjl{{\it Ap.\ J.\ Lett.}}
\newcommand\araa{{\it Ann. Rev. Astron. Astroph.}}
\newcommand\physrep{{\it Phys. Rept.}}
\newcommand\apjs{{\it Ap.\ J.\ Suppl.\ }}
\newcommand\jcap{{\it JCAP }}
\newcommand\prd{{\it Phys.\ Rev.\ }{\bf D}\ }
\newcommand\aap{{\it A \& A\ }}

\newcommand{\cobe}{{\sl COBE}}
\newcommand{\wmap}{{\sl WMAP}}
\newcommand{\planck}{{\sl Planck}}

\setlength{\headheight}{14pt}

% subsubsubsections and subsubsections are numbered as well as chapters, sections and subsections.
\setcounter{secnumdepth}{3}

% macros
\newcommand{\neff}{\ensuremath{N_\mathrm{eff}}}
\newcommand{\fsky}{\ensuremath{f_\mathrm{sky}}}
\newcommand{\comment}[1]{{\color{blue}{Comment: #1}}}
\newcommand{\XX}{\mathit{XX}}
\newcommand{\TT}{\mathit{TT}}
\newcommand{\TE}{\mathit{TE}}
\newcommand{\EB}{\mathit{EB}}
\newcommand{\EE}{\mathit{EE}}
\newcommand{\BB}{\mathit{BB}}
\newcommand{\TTilde}[2]{\mathit{\tilde #1\!\tilde #2}}
\newcommand{\ttilde}[2]{\mathit{\tilde #1\tilde #2}}
\newcommand{\tWX}{\TTilde{W}{X}}
\newcommand{\tST}{\TTilde{S}{T}}
\newcommand{\tYZ}{\TTilde{Y}{Z}}
\newcommand{\tWW}{\TTilde{W}{W}}
\newcommand{\tXX}{\TTilde{X}{X}}
\newcommand{\tUV}{\TTilde{U}{V}}
\newcommand{\tEE}{\TTilde{E}{E}}
\newcommand{\tEB}{\TTilde{E}{B}}
\newcommand{\tBB}{\TTilde{B}{B}}
\newcommand{\expt}{\mathrm{expt}}
\newcommand{\eq}[1]{(\ref{eq:#1})} 
\newcommand{\eqq}[1]{Eq.~(\ref{eq:#1})} 
\definecolor{orange}{rgb}{1,0.3,0}
\newcommand\comments[1]{\textcolor{orange}{[#1]} }


% cut and pasted from Neff chapter
\def\Neff{N_{\rm eff}}
\def\Nf{N_{\rm eff}}
\def\gs{g_{\star}}
\def\Mpl{M_{\rm P}}
%\def\gtrsim{\raise-.75ex\hbox{$\buildrel>\over\sim$}}
\def\lsim{\raise-.75ex\hbox{$\buildrel<\over\sim$}}


\DeclareUrlCommand\email{\urlstyle{rm}}

\begin{document}

%\title{CMB-S4 Science Book \\ Working Draft}
%\author{CMB-S4 Collaboration}
%\maketitle


\def\bibname{References}

\bibliographystyle{utphys}  %%%% MODIFIED FOR CF %%%%
%\bibliographystyle{plain}

\raggedbottom

\pagenumbering{roman}

\parindent=0pt
\parskip=8pt
\setlength{\evensidemargin}{0pt}
\setlength{\oddsidemargin}{0pt}
\setlength{\marginparsep}{0.0in}
\setlength{\marginparwidth}{0.0in}
\marginparpush=0pt

% The content begins here

\renewcommand{\chapname}{chap:intro_}
\renewcommand{\chapterdir}{.}
\renewcommand{\arraystretch}{1.25}
\addtolength{\arraycolsep}{-3pt}

\pagenumbering{roman} 
\chapter*{CMB-S4 Science Book\\ First Edition}
\vskip -9.5pt
\hbox to\headwidth{%
       \leaders\hrule height1.5pt\hfil}
\vskip-6.5pt
\hbox to\headwidth{%
       \leaders\hrule height3.5pt\hfil}


  \begin{center}
   {\Large\bf
      CMB-S4 Collaboration\\
      \bigskip
      August 1, 2016
   }
   \vskip 1 in
   {\Large\bf
   This advanced copy is being provided prior to posting with the list of contributors on the public archive.
   
   % Draft date: \today
%  March 21, 2016
   }
 \end{center}
\eject

\setcounter{page}{1}

\input ExecSum/exec-sum.tex

\begin{center}
  {\Large \bf Preface}
\end{center}
\bigskip

This Science Book is the product of a large, global community of scientists who are united in support of proceeding with CMB-S4, which will make key advances in our understanding of the fundamental nature of space and time and the evolution of the Universe.
%This Science Book is the product of a large community of scientists spanning the globe that are united 
%%in their enthusiasm for CMB research and 
%in their understanding of the critical need to proceed with CMB-S4 to make key advances in our understanding of the fundamental nature of space and time and the evolution of the Universe.  
The CMB-S4 concept was conceived during the 2013 Snowmass Cosmic Frontier planning exercise. Through the Snowmass process including two meetings and numerous telecons, the CMB experimental groups and the broader cosmology community came together to produce two influential CMB planning papers, endorsed by over 90 scientists, that outlined the science case as well as the CMB-S4 instrumental concept \cite{snowmass_inf2015,snowmass_neu2015}.  It became clear that an enormous increase in the scale of ground-based CMB experiments would be needed to achieve the exciting scientific goals, necessitating a phase change in the ground-based CMB experimental program. To realize CMB-S4, a partnership of the university-based CMB groups, the broader cosmology community and the national laboratories would be needed.

Based on the Snowmass papers and with additional information from the CMB experimental groups, the 2014 report of the Particle Physics Project Prioritization Panel (P5) included CMB-S4 in their recommended program.  After the P5 report was released, the CMB community began a series of biannual workshops to advance CMB-S4. The first of these was a dedicated session at the  {\it Cosmology with the CMB 
and its Polarization} workshop at the University of Minnesota January 14-16, 2015 attended by over 90 scientists.
% 94 registrants
Discussions focused on the unique and vital role of the future ground-based CMB program and its synergy with a possible future satellite mission. It was decided that the community would draft a detailed CMB-S4 Science Book.  The second and third workshops, {\it Cosmology with CMB-S4} held at the University of Michigan September 21-22, 2015 with over 100 participants,
%106 registrants
and {\it Cosmology with CMB-S4} held at the Lawrence Berkeley National Laboratory March 7-9, 2016 with over 160 participants,
% 160 registrants
were dedicated to developing the Science Book.  Working groups for each CMB-S4 science thrust were responsible for preparing and leading dedicated sessions at the workshop and for drafting the corresponding chapters of this book. In addition, a small writing group was responsible for integrating the Science Book. Through the workshops, numerous teleconferences, postings on the CMB-S4 wiki, contributions to the github Science Book repository, and feedback on drafts,  over 200 scientists have contributed to this first edition of the CMB-S4 Science Book.

\tableofcontents

%%%% Author comment macros here

\def\as#1{[{\bf AS:} {\it #1}] }


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%     All of your files should be in a subdirectory.  Here the
%%%     subdirectory is called CF0. The title of your
%%%     report should be wgreportCF0.tex in that subdirectory.  Input
%%%     that file here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\eject
\pagenumbering{arabic} 
\setcounter{page}{1}
\input Intro/intro_cmbs4.tex
\input Inflation/inflation_cmbs4.tex 
\input Neutrinos/neutrinos_cmbs4.tex 
\input Neff/neff_cmbs4.tex
\input DarkMatter/darkmatter_cmbs4.tex
\input DarkEnergy/darkenergy_cmbs4.tex 
\input CMBLensing/cmblensing_cmbs4.tex 
\input Analysis/analysis_cmbs4.tex 
%\input Instrumentation/instrumentation_cmbs4.tex 

\bibliography{cmbs4}

%\input conclusions_cmbs4.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%   Your subdirectory (here CF0) should include
%%%    the files:
%%%           wgreportCF0.tex
%%%           authorlistCF0.tex
%%%           bibCF0.bib
%%%         and all needed figures in pdf format
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}


